---
title: "GO.ddb overview"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{GO.ddb overview}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

We aim to provide a modern interface to Gene Ontology.
We use the Semantic SQL serialization in SQLite produced
by the INCAtools project.

# Basic setup

Tidyverse practices are supported.  Within the database,
the GO CURIE is in the `id` column.
```{r lk1, message=FALSE}
library(GO.ddb)
library(dplyr)
go_terms()
```

# Lookups

```{r dolook}
go_terms() |> head() |> collect() |> names()
go_terms() |> filter(label %like% "%apopto%")
go_terms() |> filter(id %in% c("GO:0042267", "GO:0042268", 
     "GO:0042269", "GO:0042270")) |> select(id, ontology, label)
```

A shortcut for term lookup:
```{r docur}
lookup_curie(c("GO:0006953", "GO:0001795"))
```

# Traversals of GO

## Ancestor (generalization) retrieval

`go_ancestors` utilizes the pre-computed 'entailed_edge' table
in the Semantic SQL representation of GO.

```{r doanc}
go_ancestors("GO:0006954", relations = GO_RELATIONS["is_a"]) |>
  inner_join(
    go_terms() |> select(ancestor_id = id, label),
    by = "ancestor_id"
  )
```

## Descendant

```{r dodesc}
go_descendants("GO:0006954", relations = GO_RELATIONS["is_a"]) |>
  inner_join(
    go_terms() |> select(descendant_id = id, label),
    by = "descendant_id"
  )
```

# Cross-ontology activities

## Quick overview of RDF and ontological entailments

Briefly, the fundamental representation of Gene Ontology
is RDF (Resource Description Format).  Each ontological
assertion (statement) has the form of an ordered triple (subject, predicate,
object).

All the implications of statements in Gene Ontology are
available in the Semantic SQL representation.  "Implication"
here refers to derivability in a logical system implemented
in an OWL reasoner.  See [OAK (ontology access kit)](https://incatools.github.io/ontology-access-kit/guide/relationships-and-graphs.html#further-notes-on-owl-and-graph-projection)
for more background.

## Common predicates

To express certain concepts *precisely*, references to external ontologies
are employed.  

```{r lkee}
go_entailed_edges() |> select(predicate) |> 
   group_by(predicate) |> summarise(n=n())
```

## Decoding external predicate CURIEs

We can use ontoProc2 to help resolve details
of predicates used to state relationships beyond those
related to class membership ("is a", "part of").

### Learning about an ontology

The following code probes the first 20 statements
available in the Semantic SQL representation of BSPO.
There is in principle no guarantee that such a search
would yield useful records.

```{r trybs, message=FALSE}
library(ontoProc2)
library(DT)
bspo_ref = semsql_connect(ontology="bspo")
bspo_ref
tbl(bspo_ref@con, "statements") |> head(20) |>
   collect() |> datatable()
```

### Statements related to a selected CURIE 

In this example we retrieve all statements with *subject*
"BSPO:0000098", which occurs with some frequency as a *predicate* in the
entailments of GO.

```{r trybs2, message=FALSE}
library(ontoProc2)
library(DT)
bspo_ref = semsql_connect(ontology="bspo")
bspo_ref
tbl(bspo_ref@con, "statements") |> filter(subject=="BSPO:0000098")  |> 
   collect() |> datatable()
```
